---
title: "Power Calculations Mendel-UPenn"
date: "December 16, 2024"
mainfont: Times New Roman
output:
  html_document:
    toc: true
  pdf_document:
    toc: true
geometry: margin=1in
fontsize: 11pt
---

## Study design

The primary purpose of the Mendel.ai study is to establish whether the predictions on oncology clinical trial eligibility made by a human+AI collaboration are non-inferior to the predictions made by humans alone. The predictions for trial eligibility are made by assessing eligibility across multiple clinical categories (ex. tumor stage, cancer biomarkers). 

*Primary Outcome*

The primary outcome for this study is the proportion of categories correctly classified with respect to its eligibility status. This outcome is measured for each patient. 

*Hypothesis* 

Non-inferiority is established by comparing the differences in the mean of the primary outcome between the two comparison arms with a pre-defined non-inferiority margin, $\Delta$. 

To this end, the Mendel trial tests the following one-sided hypotheses: 

$H_0: \mu_a - \mu_h \leq -\Delta$  (Null hypothesis)

$H_a: \mu_a - \mu_h > -\Delta$ (Alternative hypothesis)

| Parameter | Description | 
|:---------|:------:|
| $\mu_a$  | Mean proportion of categories correctly classified by human+ai | 
| $\mu_h$  | Mean proportion of categories correctly classified by human+ai | 
| $\Delta$ > 0 | Non-inferiority margin, defined by researcher (ex. .05, .10, .15)

## Method

The purpose of these calculations is to establish a minimum sample size required for a well-powered non-inferiority study. 

### Mean proportion correct

Let $Y_i$ represent the proportion of correctly classified criteria for the ith patient, and $S_{ij}$ denote the classification accuracy result of the jth (j=1..m) criterion for the ith (i=1..n) patient. Then, 

$S_{ij}$ ~ Bernoulli(E($S_{ij}$)), where $E(S_{ij}) = P(S_{ij} = 1)$ = probability recorded observation is correct for category j 

$S_{ij} =
  \begin{cases}
    1     & \text{if correct}, \\
    0 & \text{if not correct}.
  \end{cases}$

Then $Y_i =  \sum_{j=1}^{m} \frac{S_{ij}}{m}$ 

Let $Z_{ij}$ denote the classification result of the jth criterion for the ith patient, where

$Z_{ij}$ ~ Bernoulli(E($Z_{ij}$)), where $E(Z_{ij}) = P(Z_{ij} = 1)$ = probability recorded observation is marked eligible for category j

$Z_{ij} =
  \begin{cases}
    1     & \text{if marked eligible}, \\
    0 & \text{if marked not eligible}.
  \end{cases}$

And let $T_{ij}$ denote the true classification of the jth criterion for the ith patient, where

$T_{ij}$ ~ Bernoulli(E($T_{ij}$)), where $E(T_{ij}) = P(T_{ij} = 1)$ = probability recorded observation is truly eligible for category j (event rate)

$T_{ij} =
  \begin{cases}
    1     & \text{if truly eligible}, \\
    0 & \text{if truly not eligible}.
  \end{cases}$


Then: 

$\mu = E(Y_i) = E(\sum_{j=1}^{m} \frac{S_{ij}}{m}) = \frac{1}{m}\sum_{j=1}^{m}E(S_{ij}) = \frac{1}{m}\sum_{j=1}^{m}P(S_{ij}=1)$ 

where 

$P(S_{ij} = 1) = P(Z_{ij} = 1, T_{ij} = 1) + P(Z_{ij} = 0, T_{ij} = 0)$

= P($Z_{ij}$ = 1 | $T_{ij}$ = 1) P($T_{ij}$ = 1) + P($Z_{ij}$ = 0 | $T_{ij}$ = 0) P($T_{ij}$ = 0) = $TPR_j*p_j + (1-FPR_j)*(1-p_j) = TPR_j*p_j + TNR_j*(1-p_j)$

=$\frac{TP_j+TN_j}{TP_j+TN_j+FP_j+FN_j}$

and 

$Var(Y_{i}) = Var(\frac{1}{m}\sum_jS_{ij}) = \frac{1}{m^2}\sum_jVar(S_{ij}) = \frac{1}{m^2}\sum_j[P(S_{ij} = 1)(1-P(S_{ij} = 1))]$

We calculate $\mu$, a constant mean proportion of correctly classified criterion, for both the human+ai collaboration arm, $\mu_a$ = E($Y_i^a$), and the human arm, $\mu_h$ = E($Y_i^h$). We assume independence in the evaluation of different individuals and different categories.

__Code for finding vector of probability of success P(S_ij = 1), expectation of Y_ij, and variance of Y_ij__

```{r}
p.s <- function(TPR, FPR, event_rate){TPR*event_rate + (1-FPR)*(1-event_rate)} #finding probability of success/correct for category j if TPR, FPR are given for category j
mu <- function(p.s_vector){(1/length(p.s_vector))*sum(p.s_vector)} #input is the vector of probabilities of correct for all categories j=1..m
var <- function(p.s_vector){(1/(length(p.s_vector))^2)*sum(p.s_vector*(1-p.s_vector))}#input is the vector of probabilities of correct for all categories j=1..m
```

### Test statistic

When the Human and Human+AI arms consist of different group of patients, the calculation of sample size can be performed using standard formula for testing equivalence of two proportions. Assuming both arms recruit n patients, the test statistic is formed by:

$T_n$= $\frac{(\mu_a -\mu_h) - (-\Delta)}{\sqrt{Var(D)/n}}$ = $\frac{(\mu_a -\mu_h) + \Delta}{\sqrt{Var(D)/n}}$, where D = $\mu_a -\mu_h$ and Var(D) are calculated below:

__Paired design__

In this study design, the human and human+AI collaboration arms have the same sample population leading to paired data.

D = $\frac{1}{n}\sum_{i=1}^{n}Y_{i}^a-Y_{i}^h$ 

Var(D) = $Var(\frac{1}{n}\sum_{i=1}^{n}Y_{i}^a-Y_{i}^h) = \frac{1}{n^2}\sum_iVar(Y_{i}^a-Y_{i}^h) = \frac{1}{n}[Var(Y_i^a ) + Var(Y_i^h) - 2Cov(Y_i^a, Y_i^h)]$

Let $Var(D^{'}) = Var(Y_i^a ) + Var(Y_i^h) - 2Cov(Y_i^a, Y_i^h)$, where 

$Cov(Y_i^a, Y_i^h) = \rho\sqrt{Var(Y_i^a)Var(Y_i^h)} = E(Y_i^aY_i^h) - E(Y_i^a)E(Y_i^h) = E(Y_i^aY_i^h) - \mu_a\mu_h$

Alternative approach, concordance rate different between criterion: 

$Cov(Y_i^a, Y_i^h) = Cov(\frac{1}{m}\sum_jS_{ij}^a, \frac{1}{m}\sum_jS_{ij}^h) = \frac{1}{m^2}\sum_jCov(S_{ij}^a, S_{ij}^h)$, where we assume independence between individuals and between criterion and 

$Cov(S_{ij}^a, S_{ij}^h) = E(S_{ij}^aS_{ij}^h) - E(S_{ij}^a)E(S_{ij}^h) = P(S_{ij}^a = 1, S_{ij}^h=1) - P(S_{ij}^a = 1)P(S_{ij}^h = 1)$, where

$P(S_{ij}^a = 1, S_{ij}^h=1)$ = $P(Z_{ij}^a = 1, Z_{ij}^h=1, T_{ij} =1) + P(Z_{ij}^a = 0, Z_{ij}^h=0, T_{ij} =0)$

= $P(Z_{ij}^a = 1| Z_{ij}^h=1, T_{ij} =1)P(Z_{ij}^h=1 | T_{ij} =1)P(T_{ij} =1) + P(Z_{ij}^a =0| Z_{ij}^h=0, T_{ij} =0)P(Z_{ij}^h=0 | T_{ij} =0)P(T_{ij} =0)$

=$P(Z_{ij}^a = 1| Z_{ij}^h=1, T_{ij} =1)TPR_j^hp_j + P(Z_{ij}^a = 0| Z_{ij}^h=0, T_{ij} =0)TNR_j^h(1-p_j)$

We may assume $P(Z_{ij}^a = 1| Z_{ij}^h=1, T_{ij} =1)$ = $P(Z_{ij}^a = 0| Z_{ij}^h=0, T_{ij} =0)$ = 1

Then,

$Cov(Y_i^a, Y_i^h) = \frac{1}{m^2}\sum_jCov(S_{ij}^a, S_{ij}^h) = \frac{1}{m^2}\sum_j[P(S_{ij}^a = 1, S_{ij}^h=1) - P(S_{ij}^a = 1)P(S_{ij}^h = 1)]$

__Code to get Var(D')__

```{r}
#p11 represents P(Z_ij^a=1 | Z_ij^h=1, T_ij = 1) = probability AI + human collaboration is correct for observation i if human was correct for that observation
#p00 represents P(Z_ij^a-0 | Z_ij^h=0, T_ij = 1) = probability AI + human collaboration is incorrect for observation i if human was incorrect for that observation 
var_paired <- function(p11, p00, p.s_vector_1, p.s_vector_0, event_rate,fpr_0, tpr_0){ # if probability of concordance is known and provided 
  cov <- (1/length(p.s_vector_1))^2*sum((p11*tpr_0*event_rate + p00*(1-fpr_0)*(1-event_rate)) - (p.s_vector_0*p.s_vector_1))
  return(var(p.s_vector_1) + var(p.s_vector_0) - 2*cov)
}
```

### Calculating sample size for a paired design 

n = $\frac{(Z_{1-\beta} + Z_{1-\alpha})^2Var(D^{'})}{((\mu_a - \mu_h) +\Delta)^2}$, where  $\beta$ is Type II error, $1-\beta$ is the desired power, $\alpha$ is Type-I error, and $Var(D^{'})$ is calculated for paired or unpaired designs as specified above

```{r}
find_n <- function(tpr_h, tpr_a, fpr_h, fpr_a, event_rate, p11, p00, delta,power, alpha, paired = TRUE){ #r = allocation ratio ; delta = margin of noninferiority (delta > 0)
  #calculate probability of success S_ij for human and human/ai arms
  p.s_h <- p.s(tpr_h, fpr_h, event_rate)
  p.s_a <- p.s(tpr_a, fpr_a, event_rate)
  
  #calculating Y_ij expectation and variance for human and human/ai arms
  mu_h <- mu(p.s_h)
  mu_a <- mu(p.s_a)
  
  var_h <- var(p.s_h)
  var_a <- var(p.s_a)
  
  #calculating variance of test statistic
  if(paired == TRUE ){
    var_d_prime <- var_paired(p11, p00, p.s_a, p.s_h, event_rate, fpr_h, tpr_h)
  }
  else{
    var_d_prime <- var_h + var_a #can also use the function var_unpaired above to calculate
  }
  
  #calculating sample size
  delta_adjusted <- delta*(mu_h+mu_a)/2 
  numerator_n <- ((qnorm(power) + qnorm(1-alpha))^2)*var_d_prime
  denominator_n <- ((mu_a - mu_h) + delta_adjusted)^2
  n <- numerator_n/denominator_n
  
  return(n)
}
```

### Calculating n with mendel data

__Loading and processing data__
```{r}
library(ggplot2)
data <- read.csv("~/Desktop/Desktop - ceb-bios-363/PHD/Mendel_Sample_Size/mendel_2.csv")
names(data)[6] <- "tpr_a"
names(data)[9] <- "event_rate"
data$tnr_h <- 1 #assumption we are making
data$tnr_a <- 1 #assumption
data$fpr_h <- 1-data$tnr_h
data$fpr_a <- 1-data$tnr_a
print(data)
```


__Paired design, changing event rate, concordance, and tnr__

Power = .8

alpha = .05

delta = .05

$P(Z_{ij}^a = 1| Z_{ij}^h=1, T_{ij} =1)$ = $P(Z_{ij}^a = 0| Z_{ij}^h=0, T_{ij} =0)$ = 0.0, 0.2, 0.4, 0.6, 0.8, 1.0

TNR = .4, .6, .8, 1.0

event_rate = 0.10, 0.20,.3, 0.40, .5,  0.60, .7, 0.80, .9, 1.0

```{r, echo = FALSE}
power = .8
alpha = .05
delta = .05
tnr <- c(.2, .4, .6, .8, 1.0)
event_rate <- seq(.1,1,.1)
p11 = p00 = seq(0,1,.2)

coord <- expand.grid(delta, power, alpha, tnr, event_rate, p11)

ss <- c()
for(i in 1:nrow(coord)){
  ss <- c(ss, find_n(data$tpr_h, data$tpr_a, 1-coord[i,4], 1-coord[i,4], coord[i,5], coord[i,6], coord[i,6], coord[i,1], coord[i,2], coord[i,3], paired = TRUE))
}

results <- data.frame(coord, ss)
names(results) <- c("delta", "power", "alpha", "tnr", "event_rate", "p11", "n")
results$p11 <- as.character(results$p11)
results$tnr <- as.character(results$tnr)

pdf("~/Desktop/Desktop - ceb-bios-363/PHD/Mendel_Sample_Size/20241114_Mendel_Paired_Sample_Size.pdf", width = 10, height = 7)
ggplot(results, aes(x = event_rate, y = n, group = p11, color = p11)) + geom_line() + facet_wrap(~tnr) + theme_bw() + scale_y_continuous(breaks = round(seq(0, 1200, by = 50),1)) +
  labs(x= "Event Rate", y =  "Minimum Sample Size Required") + theme(axis.title = element_text(size=10, face = "bold"), axis.text=element_text(size = 10, face ="bold"), axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, face = "bold"),axis.text.y = element_text(angle = 0, hjust = 0.5, vjust = 0.5, face = "bold"), axis.line = element_line(size = 1), axis.ticks = element_line(size = 0.5), panel.grid = element_blank())
dev.off()

ggplot(results, aes(x = event_rate, y = n, group = p11, color = p11)) + geom_line() + facet_wrap(~tnr) + theme_bw() + scale_y_continuous(breaks = round(seq(0, 1200, by = 50),1)) +
  labs(x= "Event Rate", y =  "Minimum Sample Size Required") + theme(axis.title = element_text(size=10, face = "bold"), axis.text=element_text(size = 10, face ="bold"), axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 0.5, face = "bold"),axis.text.y = element_text(angle = 0, hjust = 0.5, vjust = 0.5, face = "bold"), axis.line = element_line(size = 1), axis.ticks = element_line(size = 0.5), panel.grid = element_blank())
```









